\begin{flushleft} \textsf{\small Statistics and Computing} \end{flushleft} \vspace{0.5cm} Supplementary material for "Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming"

### 2020-09-10 Paper/online_supplemental_material/online_material.tex
update supplemental material
updating supplementary material

\documentclass[onecolumn,a4paper,11pt]{article}
\voffset -1cm
\hoffset -1cm
\textheight 22.3cm
\textwidth 15.0cm

@@

\usepackage{natbib}
\bibliographystyle{plainnat}
\bibpunct{(}{)}{;}{a}{,}{,}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{float}
\usepackage{soul}
\usepackage{xcolor}
\graphicspath{{../manuscript_v2/images/}}

@@

\usepackage{listings}
\usepackage{slashbox}
\usepackage{caption}
\captionsetup{font=footnotesize}
\usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black]{hyperref}

@@

\title{Online supplemental material of the article:\\ \vspace{7mm} 'Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming'\\ \vspace{7mm}
\normalsize by}

\author{Gabriel Riutort-Mayol and Paul-Christian Bürkner and Michael R.\ Andersen and\\
  Arno Solin and Aki Vehtari}

\date{}

@@

This document consists of four case studies of application of the article 'Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming'.

@@

\noindent Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $\bm{K}$, where $K_{ij}=k(x_i,x_j,\theta)$, with $i,j=1,\dots,74$. The covariance function $k$ depends on the inputs $\bm{x}$ and hyperparameters $\theta=\{\alpha,\ell\}$. The hyperparameters $\alpha$ and $\ell$ represent the marginal variance and lengthscale, respectively, of the GP process.

@@

In order to do model comparison, in addition to the regular GP model and HSGP model, an splines-based model is also fitted using the Thin Plate Regression Splines approach in \cite{wood2003thin} and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this splines model using the R-package \textit{brms} \citep{burkner2017brms}.

Figure \ref{ch5_fig12_Posteriors_gaydata} shows the posterior mean predictive distributions of the three models, the regular GP, the HSGP model with $m=20$ basis functions and boundary factor $c=1.5$, and the splines model with 20 knots. Sample observations are plotted as circles in the figure, and the out-of-sample observations, which have been used for testing, are plotted as crosses.

\begin{figure}
\centering
\includegraphics[scale=0.50]{ch5_fig12_Posteriors_gaydata.pdf}
\caption{Posterior predictive means of the proposed HSGP model, the regular GP model, and the Splines model. 95\% confident intervals are plotted as dashed lines.}
  \label{ch5_fig12_Posteriors_gaydata}
\end{figure}

For the HSGP model, different models with different number of basis functions and boundary factor have been fitted. The root mean square errors (RMSE) for every one of these models have been computed against the regular GP model, and plotted as a function of the number of basis functions $m$ and boundary factor $c$ in Figure \ref{ch5_fig13_MSE_train_BF_gaydata}-left, for sample data, and in Figure \ref{ch5_fig13_MSE_train_BF_gaydata}-right, for test data. As the boundary factor increases, more basis functions are needed for a accurate approximation.

Figure \ref{ch5_fig14_MSE_train_gaydata} shows the RMSE of the regular GP, HSGP and splines models computed against the actual data, for training and test data, as a function of the number of basis functions $m$ and boundary factor $c$ for the HSGP model, and  knots for the splines model. We can see how the splines models do not extrapolate data properly (Figure \ref{ch5_fig14_MSE_train_gaydata}-right).

Figure \ref{ch5_fig14_1_time_gaydata} shows computational times, in seconds per iteration (iteration of the HMC sampling method), as a function of the number of basis functions $m$, for the HSGP model, and knots, for the splines model. The computational times is represented in the y-axis which is on a logarithmic scale. The HSGP model is on average roughly 15 times faster than the regular GP and 5 times faster than the spline model, for this particular case and univariate input space. Computation time increases relatively slight with the number of basis functions in a univariate input space, as can be seen in the figure.

The Stan model codes for the exact GP, the approximate GP and the splines models of this case study can be found online at {\small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_Same-sex-marriage-data}}\,.

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig13_MSE_train_BF_gaydata.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig13_MSE_pred_BF_gaydata.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 28mm 45mm 119mm 0mm, clip]{ch5_fig13_legend_gaydata.pdf}}
\caption{Root mean square error (RMSE) of the HSGP model, computed against the regular GP model, as a function of the number of basis functions $m$ and boundary factor $c$. RMSE for sample data (left) and RMSE for out-of-sample data (right).}
  \label{ch5_fig13_MSE_train_BF_gaydata}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig14_MSE_train_gaydata.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig14_MSE_pred_gaydata.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 27mm 35mm 112mm 0mm, clip]{ch5_fig14_legend_gaydata.pdf}}
\caption{Root mean square error (RMSE) of the different methods, regular GP, HSGP and splines models, computed against the actual data, as a function of the number of basis functions $m$ and boundary factor $c$ for the HSGP model, and knots for the splines model. RMSE for sample data (left) and RMSE for out-of-sample data (right).}
  \label{ch5_fig14_MSE_train_gaydata}
\end{figure}

@@

This example consists of a simulated dataset with $n=120$ ($i=1,\dots,n$) single draws from a GP prior as function of input values $\bm{x}_i \in \{[-1,1],[-1,1]\} \subset {\rm I\!R}^2$ in a bivariate input space ($D=2$). A squared exponential covariance function, with hyperparameters marginal variance $\alpha=1$ and lengthscales $\ell_1=0.10$, for the first input dimension, and $\ell_2=0.35$, for the second input dimension, is used for the GP prior. Gaussian noise standard deviation $\sigma=0.2$ was added to the GP draws to form the final noisy set of observations $\bm{y} \in {\rm I\!R}^{120}$.

The regular GP model over the outcome variable $\bm{y}$ can be written as follows,
%
\begin{eqnarray*}\label{ch5_eq_latentgp_simudata2}
\begin{split}
\bm{y} &= \bm{f} + \bm{\epsilon} \\
\bm{\epsilon} &\sim \Normal(0, \sigma^2  I) \\
f(\bm{x}) &\sim \mathcal{GP}(0, k(\bm{x}, \bm{x}', \theta)),
\end{split}
\end{eqnarray*}

\noindent where $\bm{f}=\{f(\bm{x}_i)\}_{i=1}^{120}$ represents the underlying function at the input values $\bm{x_i}\in {\rm I\!R}^{2}$, $\bm{\epsilon}$ is the Gaussian noise term with variance $\sigma^2$, and $I$ represents the identity matrix. The function $f:{\rm I\!R}^2 \to {\rm I\!R}$ is a GP prior with a multivariate squared exponential covariance function $k$, which depends on the inputs $\bm{x}$ and hyperparameters $\theta=\{\alpha,\ell_1,\ell_2\}$. The hyperparameters $\alpha$, $\ell_1$ and $\ell_2$ represent the marginal variance and lengthscales for first and second input dimensions, respectively, of the GP process.

@@

In the HSGP model with $2$ input dimensions, the latent function $f(\bm{x})$ is approximated as in equation (12)% (\ref{eq_approxf_multi})
, with the $2$-dimensional squared exponential spectral density $S$ as in equation (1)% (\ref{eq_specdens_inf})
, and the $2$-vector of eigenvalues $\bm{\lambda}_j$ and the multivariate eigenfunctions $\phi_j$ as in equations (9) %(\ref{eq_eigenfunction_multi})
 and (10)% (\ref{eq_eigenvalue_multi})
 , respectively.

In order to do model comparison, in addition to the regular GP and HSGP models, a two-dimensional splines-based model is also fitted using a cubic spline basis, penalized by the conventional integrated square second derivative cubic spline penalty \citep{wood2017generalized}, and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline model using the R-package \textit{brms} \citep{burkner2017brms}.

Figure \ref{ch5_fig15_Posteriors_exII} shows the data-generating GP function, from where the dataset was drawn, and the mean posterior predictive functions of the three models, regular GP, HSGP, and splines, fitted over the dataset. Sample observations are also plotted as circles. For the HSGP model, $m_1=40$ and $m_2=40$ basis functions for each dimension respectively, were used, which lead to a total of 1600 multivariate basis functions. A boundary factor for each dimension $c_1=1.5$ and $c_2=1.5$ were used. For the splines model, 40 knots in each dimension were used.

Figure \ref{ch5_fig16_errors_exII} shows the difference functions between the data-generating function and the GP, HSGP and splines models, respectively.

@@

In order to assess performance of the models as a function of the number of basis functions and knots, different models with different number of basis functions, for the HSGP model, and different number of knots, for the splines model, have been fitted. In all models, the same number of basis functions and knots per dimension were used. Figure \ref{ch5_fig17_RMSE_exII}-left shows the root mean squared error (RMSE), computed against the data-generating function, as a function of the boundary factor $c$, and the number of univariate basis functions $m$, for the HSGP model, and knots, for the splines model. From Figures \ref{ch5_fig16_errors_exII} and \ref{ch5_fig17_RMSE_exII}-left, it can be seen a close approximation of the HSGP model to the regular GP model. However, the performance of the splines model is significantly worse.

Figure \ref{ch5_fig17_RMSE_exII}-right shows the computational times of the different models as a function of the boundary factor, number of basis functions and knots. Figure \ref{ch5_fig17_RMSE_exII} reveals that choosing the optimal boundary factor allows for less number of basis functions and less computation time. Even though in a bivariate input space the computation demand increases significantly with the number of dimensions or knots, either the HSGP or spline models work significantly better than regular GP, even for highly wiggly functions that require a high number of basis functions or knots for a accurate approximation. However, serious difficulties with cumputation time were encountered in building the spline model with $50$ knots for this bivariate function.

The Stan model codes for the exact GP, the approximate GP and the splines models of this case study can be found online at {\small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_2D-Simulated-data}}\,.

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig17_RMSE_exII.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig17_time_exII.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 28mm 35mm 112mm 0mm, clip]{ch5_fig17_legend_exII.pdf}}
\caption{Root mean square error (RMSE) (left) and computational time (right) in seconds per iteration (iteration of the HMC sampling method) of the different methods computed against the data-generating function, as a function of the boundary factor $c$, number of basis functions $m$ and knots.}
  \label{ch5_fig17_RMSE_exII}
\end{figure}

\section{Diabetes case study}\label{ch5_sec_bf_caseV}
The next example presents an epidemiological study of diabetes disease. The study aims to relate the probability of suffering from diabetes to some risk factors. The data contains $n=392$ individuals ($i=1,\dots,n$) from which the binary variable of suffering ($y_i=1$) or not suffering ($y_i=0$) from diabetes have been observed. The input vector $\bm{x}_i=(x_{i1},x_{i2},x_{i3},x_{i4}) \in {\rm I\!R}^{4}$, in a 4D input space ($D=4$), contains the risk factors, \textit{Glucose} ($x_{i1}$), \textit{Pregnancy} ($x_{i2}$), \textit{Age} ($x_{i3}$) and \textit{BMI} ($x_{i4}$), per individual $i$. The observational model is a Bernoulli model with parameter the probability $p_i$ of suffering from diabetes per observation $i$,
%
\begin{equation*}
y_i \sim \mathrm{Bernoulli}(p_i).
\end{equation*}

\noindent The goal is to estimate the probability $p_i$ as a function of the risk factors, which function $f(\cdot):{\rm I\!R}^{4} \to {\rm I\!R}$ is modeled as a GP with a multivariate squared exponential covariance function $k$ depending on the risk factors $\bm{x}$ and hyperparameters $\theta=\{\alpha,\ell\}$, and related to the probabilities $p_i$ through the {\it logit} link function,
%
\begin{align*} \label{ch5_eq_gpprior_gay}
p_i &= \mathrm{logit}(f(\bm{x}_i)) \nonumber \\
f(\bm{x}) &\sim \mathcal{GP}(0, k(\bm{x},\bm{x}', \theta).
\end{align*}

\noindent The hyperparameters $\alpha$ and $\ell$ represent the marginal variance and lengthscale, respectively, of the GP process. A scalar lengthscale has been considered in the multivariate covariance function $k$.

In the HSGP model with $D$ input dimensions, the function $f(\bm{x})$ evaluated at input vector $\bm{x} \in {\rm I\!R}^D$ is approximated as in equation (12)% (\ref{eq_approxf_multi})
, with the $D$-dimensional (with a scalar lengthscale) squared exponential spectral density $S$ as in equation (1) %(\ref{eq_specdens_inf}) 
and the $D$-vector of eigenvalues $\bm{\lambda}_j$ and the multivariate eigenfunctions $\phi_j$ as in equations (9) % (\ref{eq_eigenvalue_multi}) 
and (10)% (\ref{eq_eigenfunction_multi})
, respectively.

In order to do model comparison, in addition to the regular GP and HSGP models, a $D$-dimensional spline-based model is also fitted using a cubic spline basis penalized by the conventional integrated square second derivative cubic spline penalty \citep{wood2017generalized} and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline model using the R-package \textit{brms} \citep{burkner2017brms}.

Figure \ref{ch5_fig18_gpfun_diabetes} shows the posterior mean predictions of probabilities ($p_i$) of the three models, regular GP, HSGP and spline, fitted over the dataset with the 2 input dimensions \textit{Glucose} and \textit{Pregnancy} ($D=2$). The binary observations $y_i$ are also plotted in the plots as colored points. For the HSGP model, $m_1=20$ and $m_2=20$ basis functions for each dimension, respectively, were used, which lead to a total of 400 multivariate basis functions. A boundary factor for each dimension $c_1=4$ and $c_2=4$ were used. For the spline model, 20 knots per dimension were used.
%
\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.37, trim = 5mm 23mm 39mm 20mm, clip]{ch5_fig18_gpfun_diabetes.pdf}}
\subfigure{\includegraphics[scale=0.37, trim = 20mm 23mm 39mm 20mm, clip]{ch5_fig18_bffun_diabetes.pdf}}
\subfigure{\includegraphics[scale=0.37, trim = 20mm 23mm 39mm 20mm, clip]{ch5_fig18_spfun_diabetes.pdf}}
\subfigure{\includegraphics[scale=0.30, trim = 150mm 4mm 2mm 15mm, clip]{ch5_fig18_legend_diabetes.pdf}}
\caption{Posterior predictive mean functions of the GP (a), HSGP (b) and spline (c) models. Samples observations of suffering (red points) and not suffering (blue points) from the disease are plotted.}
  \label{ch5_fig18_gpfun_diabetes}
\end{figure}

In order to assess the performance of the models as a function of the boundary factor, the number of basis functions and knots, different models with different number of basis functions and boundary factor for the HSGP model and different number of knots for the spline model have been fitted. In all models, the same boundary factor, number of basis functions and knots per dimension were used. Figure \ref{ch5_fig19_ELPD_diabetes} shows the expected log predictive density (ELPD; see \cite{vehtari_2012}) as a function of the boundary factor $c$ and the number of univariate basis functions $m$, for the HSGP model, and knots, for the spline model. The ELPD is computed over the actual observations by cross-validation. Basically, with slightly differences, all models show similar performances, due to the fact that the process is very smooth with a relatively very large lengthscale estimate $\ell=4.51$. Even though, a slight pattern of performance improvement can be appreciated as the boundary factor $c$ increases, which fact is because small boundary factors are not allowed when large lengthscales (Figure 6% \ref{fig6_lscale_vs_J_vs_c}
).
%
\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig19_ELPD_diabetes.pdf}}
\caption{Expected log predictive density (ELPD) of the different methods as a function of the boundary factor $c$ and the number of basis functions $m$, for the HSGP model, and knots, for the spline model.}
  \label{ch5_fig19_ELPD_diabetes}
\end{figure}

>>>>

\documentclass[onecolumn,a4paper,11pt]{article}
\voffset -2cm
\hoffset -1cm
\textheight 23.3cm
\textwidth 15.0cm

@@

\usepackage{natbib}
\bibliographystyle{plainnat}
\bibpunct{(}{)}{;}{a}{,}{,}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{float}
\usepackage{soul}
\usepackage{xcolor}
\graphicspath{{./images/}}

@@

\usepackage{listings}
\usepackage{slashbox}
\usepackage{caption}
\captionsetup{font=footnotesize}
\usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,breaklinks=true]{hyperref}

@@

\title{\begin{flushleft}
\textsf{\small Statistics and Computing}
\end{flushleft}  \vspace{0.5cm}  Supplementary material for "Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming"}

\author{Gabriel Riutort-Mayol$^{1*}$, Paul-Christian Bürkner$^2$, Michael R.\ Andersen$^{3}$,\\
  Arno Solin$^{2}$, Aki Vehtari$^{2}$}

\date{ \small
$^1$ Department of Cartographic Engineering, Geodesy, and Photogrammetry, Universitat Polit\`ecnica de Val\`encia, Spain 
\break
$^2$ Department of Computer Science, Aalto University, Finland
\break
$^3$ Department of Applied Mathematics and Computer Science, Technical University of Denmark, Denmark
\break
$^*$ Corresponding author, Email: gabriuma@gmail.com
}

@@

This supplementary material consists of four case studies of application of the methodology proposed in the article.

@@

\noindent Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $\bm{K}$, where $K_{ij}=k(x_i,x_j,\theta)$, with $i,j=1,\dots,74$. The covariance function $k$ depends on the inputs $\bm{x}$ and hyperparameters $\theta=\{\alpha,\ell\}$. The hyperparameters $\alpha$ and $\ell$ represent the marginal variance and length-scale, respectively, of the GP process.

@@

In order to do model comparison, in addition to the exact GP model and HSGP model, an spline-based model is also fitted using the Thin Plate Regression Splines approach in \cite{wood2003thin} and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline-based model using the R-package \textit{brms} \citep{burkner2017brms}.

Figure \ref{ch5_fig12_Posteriors_gaydata} shows the posterior mean predictive distributions of the three models, the exact GP, the HSGP model with $m=20$ basis functions and boundary factor $c=1.5$, and the spline-based model with 20 knots. Sample observations are plotted as circles in the figure, and the out-of-sample observations, which have been used for testing, are plotted as crosses.

\begin{figure}
\centering
\includegraphics[scale=0.50]{ch5_fig12_Posteriors_gaydata.pdf}
\caption{Posterior predictive means of the proposed HSGP model, the exact GP model, and the Splines model. 95\% confident intervals are plotted as dashed lines.}
  \label{ch5_fig12_Posteriors_gaydata}
\end{figure}

For the HSGP model, different models with different number of basis functions and boundary factor have been fitted. The root mean square errors (RMSE) for every one of these models have been computed against the exact GP model, and plotted as a function of the number of basis functions $m$ and boundary factor $c$ in Figure \ref{ch5_fig13_MSE_train_BF_gaydata}, for sample (left) and test (right) data. The expected patterns of the approximation as a function of the number of basis functions and boundary factor are recognized: as the boundary factor increases, more basis functions are needed.

Figure \ref{ch5_fig14_MSE_train_gaydata} shows the RMSE of the exact GP, HSGP and spline-based models computed against the actual data, for training and test data, as a function of the number of basis functions $m$ and boundary factor $c$ for the HSGP model, and  knots for the spline-based model. We can see how the spline-based model does not extrapolate data properly (Figure \ref{ch5_fig14_MSE_train_gaydata}-right).

Figure \ref{ch5_fig14_1_time_gaydata} shows computational times, in seconds per iteration (iteration of the HMC sampling method), as a function of the number of basis functions $m$, for the HSGP model, and knots, for the spline-based model. The computational times is represented in the y-axis which is on a logarithmic scale. The HSGP model is on average roughly 15 times faster than the exact GP and 5 times faster than the spline-based model, for this particular case and univariate input space. Computation time increases relatively slight with the number of basis functions in a univariate input space, as can be seen in the figure.

The Stan model codes for the exact GP, the HSGP and the spline-based models of this case study can be found online at {\small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_Same-sex-marriage-data}}\,.

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig13_MSE_train_BF_gaydata.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig13_MSE_pred_BF_gaydata.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 28mm 45mm 119mm 0mm, clip]{ch5_fig13_legend_gaydata.pdf}}
\caption{Root mean square error (RMSE) of the HSGP model, computed against the exact GP model, as a function of the number of basis functions $m$ and boundary factor $c$. RMSE for sample data (left) and RMSE for out-of-sample data (right).}
  \label{ch5_fig13_MSE_train_BF_gaydata}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig14_MSE_train_gaydata.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig14_MSE_pred_gaydata.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 27mm 35mm 112mm 0mm, clip]{ch5_fig14_legend_gaydata.pdf}}
\caption{Root mean square error (RMSE) of the different methods, exact GP, HSGP and spline-based models, computed against the actual data, as a function of the number of basis functions $m$ and boundary factor $c$ for the HSGP model, and knots, for the spline-based model. RMSE for sample data (left) and RMSE for out-of-sample data (right).}
  \label{ch5_fig14_MSE_train_gaydata}
\end{figure}

@@

This example consists of a synthetic dataset with $n=120$ ($i=1,\dots,n$) single draws from a GP prior as a function of input values $\bm{x}_i \in \{[-1,1],[-1,1]\} \subset {\rm I\!R}^2$ in a bivariate input space ($D=2$). A squared exponential covariance function, with hyperparameters marginal variance $\alpha=1$ and length-scales $\ell_1=0.10$, for the first input dimension, and $\ell_2=0.35$, for the second input dimension, is used for the GP prior. Gaussian noise standard deviation $\sigma=0.2$ was added to the GP draws to form the final noisy set of observations $\bm{y} \in {\rm I\!R}^{120}$.

The exact GP model over the outcome variable $\bm{y}$ can be written as follows,
%
\begin{eqnarray*}\label{ch5_eq_latentgp_simudata2}
\begin{split}
\bm{y} &= \bm{f} + \bm{\epsilon} \\
\bm{\epsilon} &\sim \Normal(0, \sigma^2  I) \\
f(\bm{x}) &\sim \mathcal{GP}(0, k(\bm{x}, \bm{x}', \theta)),
\end{split}
\end{eqnarray*}

\noindent where $\bm{f}=\{f(\bm{x}_i)\}_{i=1}^{120}$ represents the underlying function at the input values $\bm{x_i}\in {\rm I\!R}^{2}$, $\bm{\epsilon}$ is the Gaussian noise term with variance $\sigma^2$, and $I$ represents the identity matrix. The function $f:{\rm I\!R}^2 \to {\rm I\!R}$ is a GP prior with a multivariate squared exponential covariance function $k$, which depends on the inputs $\bm{x}$ and hyperparameters $\theta=\{\alpha,\ell_1,\ell_2\}$. The hyperparameters $\alpha$, $\ell_1$ and $\ell_2$ represent the marginal variance and length-scales for first and second input dimensions, respectively, of the GP process. Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $\bm{K}$, where $K_{ij}=k(\bm{x}_i,\bm{x}_j,\theta)$, with $i,j=1,\dots,120$.

@@

In the HSGP model with $2$ input dimensions, the latent function $f(\bm{x})$ is approximated as in equation (13)% (\ref{eq_approxf_multi})
, with the $2$-dimensional squared exponential spectral density $S$ as in equation (1)% (\ref{eq_specdens_inf})
, and the $2$-vector of eigenvalues $\bm{\lambda}_j$ and the multivariate eigenfunctions $\phi_j$ as in equations (10) %(\ref{eq_eigenfunction_multi})
 and (11)% (\ref{eq_eigenvalue_multi})
 , respectively.

In order to do model comparison, in addition to the exact GP and HSGP models, a two-dimensional splines-based model is also fitted using a cubic spline basis, penalized by the conventional integrated square second derivative cubic spline penalty \citep{wood2017generalized}, and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline-based model using the R-package \textit{brms} \citep{burkner2017brms}.

Figure \ref{ch5_fig15_Posteriors_exII} shows the data-generating GP function, from where the dataset was drawn, and the mean posterior predictive functions of the three models, exact GP, HSGP, and splines, fitted over the dataset. Sample observations are also plotted as circles. For the HSGP model, $m_1=40$ and $m_2=40$ basis functions for each dimension, respectively, were used, which lead to a total of 1600 multivariate basis functions. A boundary factor for each dimension, $c_1=1.5$ and $c_2=1.5$, respectively, were used. For the spline-based model, 40 knots in each dimension were used.

Figure \ref{ch5_fig16_errors_exII} shows the difference functions between the data-generating function and the GP, HSGP and spline-based models, respectively.

@@

In order to assess performance of the models as a function of the number of basis functions and knots, different models with different number of basis functions, for the HSGP model, and different number of knots, for the spline-based model, have been fitted. In each of these models, the same boundary factor, number of basis functions and knots per dimension were used. Figure \ref{ch5_fig17_RMSE_exII}-left shows the root mean squared error (RMSE), computed against the data-generating function, as a function of the boundary factor $c$ and the number of univariate basis functions $m$, for the HSGP model, and knots, for the spline-based model. From Figures \ref{ch5_fig16_errors_exII} and \ref{ch5_fig17_RMSE_exII}-left, it can be seen an accurate approximation of the HSGP model to the exact GP model. However, the performance of the spline-based model is significantly worse.

Figure \ref{ch5_fig17_RMSE_exII}-right shows the computational times of the different models as a function of the boundary factor, number of basis functions and knots. Figure \ref{ch5_fig17_RMSE_exII} reveals that choosing the optimal boundary factor allows for less number of basis functions and less computation time. Even though in a bivariate input space the computational demands increase significantly with the number of dimensions or knots, either the HSGP or spline-based models work significantly better than exact GP, even for highly wiggly functions that require a high number of basis functions or knots for an accurate approximation. It is interesting to be noticed that just the computation of the input data for the spline model in a 2D input space and considering more than 50 knots per dimension is computationally very expensive.

The Stan model codes for the exact GP, the HSGP and the spline-based models of this case study can be found online at {\small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_2D-Simulated-data}}\,.

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig17_RMSE_exII.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig17_time_exII.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 28mm 35mm 112mm 0mm, clip]{ch5_fig17_legend_exII.pdf}}
\caption{Root mean square error (RMSE) (left) and computational time (right) in seconds per iteration (iteration of the HMC sampling method) of the different methods computed against the data-generating function, and plotted as a function of the boundary factor $c$, number of basis functions $m$ and knots.}
  \label{ch5_fig17_RMSE_exII}
\end{figure}

\section{Diabetes case study}\label{ch5_sec_bf_caseV}
The next example presents an epidemiological study of diabetes disease. The study aims to relate the probability of suffering from diabetes to some risk factors. The data contains $n=392$ individuals ($i=1,\dots,n$) from which the binary variable of suffering ($y_i=1$) or not suffering ($y_i=0$) from diabetes have been observed. The input vector $\bm{x}_i=(x_{i1},x_{i2},x_{i3},x_{i4}) \in {\rm I\!R}^{4}$, in a 4D input space ($D=4$), contains the risk factors: \textit{Glucose} ($x_{i1}$), \textit{Pregnancy} ($x_{i2}$), \textit{Age} ($x_{i3}$) and \textit{BMI} ($x_{i4}$), per individual $i$. The observational model is a Bernoulli model with parameter the probability $p_i$ of suffering from diabetes per observation $i$,
%
\begin{equation*}
y_i \sim \mathrm{Bernoulli}(p_i).
\end{equation*}

\noindent The goal is to estimate the probability $p_i$ as a function $f$ of the risk factors. Which function $f(\cdot):{\rm I\!R}^{4} \to {\rm I\!R}$ is modeled as a GP with a multivariate squared exponential covariance function $k$ depending on the risk factors $\bm{x}$ and hyperparameters $\theta=\{\alpha,\ell\}$, and related to the probabilities $p_i$ through the {\it logit} link function,
%
\begin{align*} \label{ch5_eq_gpprior_gay}
p_i &= \mathrm{logit}(f(\bm{x}_i)) \nonumber \\
f(\bm{x}) &\sim \mathcal{GP}(0, k(\bm{x},\bm{x}', \theta).
\end{align*}

\noindent The hyperparameters $\alpha$ and $\ell$ represent the marginal variance and length-scale, respectively, of the GP process. A scalar length-scale has been considered in the multivariate covariance function $k$.

In the HSGP model with $D$ input dimensions, the function $f(\bm{x})$ evaluated at input vector $\bm{x} \in {\rm I\!R}^D$ is approximated as in equation (13)% (\ref{eq_approxf_multi})
, with the $D$-dimensional (with a scalar length-scale) squared exponential spectral density $S$ as in equation (1) %(\ref{eq_specdens_inf}) 
and the $D$-vector of eigenvalues $\bm{\lambda}_j$ and the multivariate eigenfunctions $\phi_j$ as in equations (10) % (\ref{eq_eigenvalue_multi}) 
and (11)% (\ref{eq_eigenfunction_multi})
, respectively.

In order to do model comparison, in addition to the exact GP and HSGP models, a $D$-dimensional spline-based model is also fitted using a cubic spline basis penalized by the conventional integrated square second derivative cubic spline penalty \citep{wood2017generalized} and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline-based model using the R-package \textit{brms} \citep{burkner2017brms}.

Figure \ref{ch5_fig18_gpfun_diabetes} shows the posterior mean predictions of probabilities ($p_i$) of the three models, exact GP, HSGP and splines, fitted over the dataset with the two input dimensions \textit{Glucose} and \textit{Pregnancy} ($D=2$). The binary observations $y_i$ are also plotted in the plots as colored points. For the HSGP model, $m_1=20$ and $m_2=20$ basis functions for each dimension, respectively, were used, which lead to a total of 400 multivariate basis functions. A boundary factor for each dimension, $c_1=4$ and $c_2=4$, respectively, were used. For the spline-based model, 20 knots per dimension were used.
%
\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.37, trim = 5mm 23mm 39mm 20mm, clip]{ch5_fig18_gpfun_diabetes.pdf}}
\subfigure{\includegraphics[scale=0.37, trim = 20mm 23mm 39mm 20mm, clip]{ch5_fig18_bffun_diabetes.pdf}}
\subfigure{\includegraphics[scale=0.37, trim = 20mm 23mm 39mm 20mm, clip]{ch5_fig18_spfun_diabetes.pdf}}
\subfigure{\includegraphics[scale=0.30, trim = 150mm 4mm 2mm 15mm, clip]{ch5_fig18_legend_diabetes.pdf}}
\caption{Posterior predictive mean functions of the GP (a), HSGP (b) and spline (c) models. Sample observations of suffering (red points) and not suffering (blue points) from the disease are plotted.}
  \label{ch5_fig18_gpfun_diabetes}
\end{figure}

In order to assess the performance of the models as a function of the boundary factor, the number of basis functions and knots, different models with different number of basis functions and boundary factors, for the HSGP model, and different number of knots, for the spline model, have been fitted. In each of these models, the same boundary factor, number of basis functions and knots per dimension were used. Figure \ref{ch5_fig19_ELPD_diabetes} shows the expected log predictive density (ELPD; see \cite{vehtari_2012}) as a function of the boundary factor $c$ and number of univariate basis functions $m$, for the HSGP model, and knots, for the spline-based model. The ELPD is computed over the actual observations by cross-validation. With slightly differences, all models show similar performances due to the fact that the process is very smooth with a relatively very large length-scale estimate $\ell=4.51$. Even though, a slight improvement can be appreciated as the boundary factor $c$ increases, because small boundary factors are not allowed when large length-scales (Figure 6% \ref{fig6_lscale_vs_J_vs_c}
).
%
\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig19_ELPD_diabetes.pdf}}
\caption{Expected log predictive density (ELPD) of the different methods as a function of the boundary factor $c$ and the number of basis functions $m$, for the HSGP model, and knots, for the spline model.}
  \label{ch5_fig19_ELPD_diabetes}
\end{figure}
